---
title: "Automated Cell Classification Workflow"
author: Givanna Putri
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
    html_document:
      toc: true
      toc_float: true
      df_print: paged
---

In this workflow, we provide an example on how to use classification to transfer the cell annotation of a dataset to another. 
This workflow is most suitable those who have performed a simple discovery workflow on a small dataset and would like to annotate a larger dataset without having to repeat the entire workflow again.
A step-by-step detailed simple discovery workflow is available on our wiki page: [Simple Discovery Workflow](https://wiki.centenary.org.au/pages/viewpage.action?pageId=151793632).

Note that if you need to perform some batch alignment to align the small *analysed and annotated* dataset and the large *to-be-annotated* dataaset, please refer to our general discovery workflow available on our wiki page: [Batch Alignment Workflow](https://wiki.centenary.org.au/pages/viewpage.action?pageId=157272386).

# Analysis setup

Here, we will simply load up the Spectre package and the demo dataset which has previously been analysed and annotated using our simple discovery workflow.

```{r}
library(Spectre)
library(data.table)

cell.dat <- demo.clustered

head(cell.dat)

```
The population is depicted in the *Population* column, and there are 6 different cell types:
```{r}
unique(cell.dat$Population)
```

# Splitting data (***FOR DEMO ONLY***)

For the purpose of demonstration, we're going to split the demo dataset into 2; the small ***annotated*** dataset and the large ***un-annotated*** dataset.
For the large dataset, we will simply remove the *Population* column.
To split the dataset, we will employ our subsample function.

*Please do not do this for your own dataset as you should already have 2 different datasets ready!*

```{r}
cell.dat$id <- c(1: nrow(cell.dat))
annot.cell.dat <- do.subsample(dat = cell.dat, targets = rep(3000, 6), divide.by = 'Population')
unannot.cell.dat <- cell.dat[!cell.dat$id %in% annot.cell.dat$id]
setnames(x = unannot.cell.dat, old = "Population", "ActualPopulation")
unannot.cell.dat$FlowSOM_cluster <- NULL
unannot.cell.dat$FlowSOM_metacluster <- NULL
unannot.cell.dat$UMAP_X <- NULL
unannot.cell.dat$UMAP_Y <- NULL
```

Let's quickly inspect the data.

Annotated dataset:
```{r}
head(annot.cell.dat)
```

Unannotated dataset, to be annotated using automated classifier:
```{r}
head(unannot.cell.dat)
```

# Running the classifier

## Determining the number of neighbours

In this workflow, we will be using the simple (yet effective!) K-Nearest Neighbour (KNN) classifier.
For those who are not familiar with KNN, here is a super brief 1 sentence explanation.
For each new un-annotated cell, KNN will identify *K* number of cells from the annotated dataset that lie the closest to it.
This is often called *neighbours*.
Then it will determine the majority population label of the neighbours and assign it to the new un-annotated cell.

Setting this *K* number is often non-trivial.
If you are not sure what is suitable, then maybe it is best to try a few and see each *K* value perform.
To do this, Spectre provides the facility to try out KNN using a range of *K* values on the annotated dataset:
```{r}
classify_cols <- c("NK11_asinh", "CD3_asinh", "CD45_asinh", "Ly6G_asinh", "CD11b_asinh", "B220_asinh", "CD8a_asinh", "Ly6C_asinh", "CD4_asinh")
knn.stats <- train.knn.classifier(dat = annot.cell.dat,
                                  use.cols = classify_cols,
                                  label.col = "Population",
                                  max.num.neighbours = 20)
```

A side note, the parameter *classify_cols* basically refer to the markers you want KNN to use when computing the distance between cells. 
Most of the time, these should be your cell type markers.

The *max.num.neighours* indicates the maximum number of neighbours the function will try. 
You can set this to any whole number > 1, but note that the larger this number is, the longer the function will take to run as it will basically run through every number between 1 to *max.num.neighbours*.
You can set a minimum number of neighbours using the parameter *min.num.neighbours*.

The message basically tells you which *K* value is being trained.

Let's inpsect the result:
```{r}
knn.stats
```

From this table, it looks like *K = 19* yielded the best accuracy score.
It is worthwhile to note that larger *K* value does not imply better performance!

## Annotating the dataset

It is as simple as running the following function:
```{r}
unannot.cell.dat <- Spectre::run.knn.classifier(train.dat = annot.cell.dat,
                                                unlabelled.dat = unannot.cell.dat,
                                                use.cols = classify_cols,
                                                label.col = "Population",
                                                num.neighbours = 19)
```

*train.dat* refers to the dataset which has previously been annotated.

*unlabelled.dat* refers to the dataset to be annotated.

*num.neighbours* refers to the number of neighbours.

Let's inspect the result:
```{r}
head(unannot.cell.dat)
```

There are several columns to be aware of:

1. Prediction: this is the cell population assigned by the KNN classifier
2. Neigbour_x: it tells you which cell in the annotated dataset is the x-th closest neighbour.

That's about it! Good luck!
